ANT+ FIT file decoder for R
===========================

```{r set-options, echo=FALSE, cache=FALSE}
library(ggplot2)
options(width = 128)
theme_set(theme_bw(20))
```

Reads FIT exercise files from a Garmin or other ANT+ devices into R. Developed for a Garmin Edge 500, but likely to work with other devices, too.

This package incorporates code from the [ANT+ FIT SDK](http://www.thisisant.com/resources/fit).

Install
-------

The easiest way to install the package is directly from Github, using Hadley Wickham's `devtools` package, as follows: 
```R
install.packages('devtools') # if you don't already have it
library(devtools)
install_github('kuperov/fit')
```
Alternatively, you can download the [R source package](https://github.com/kuperov/fit/releases/download/v0.1/fit_0.1.tar.gz) and install with `install.packages('fit_0.1.tar.gz')`.

The package depends on `Rcpp`, and there are C++ sources to compile. So you need a working C++ compiler. If you have trouble compiling the code, let me know and I'll see if I can hook you up with a binary package instead.

Getting started
---------------

The FIT protocol organises data in a series of 'messages' of different types.
Examples are exercise sessions, and laps. The fit package provides one data.frame
per message type.

First you have to extract the file from the GPS device. On my Garmin Edge 500, the files can be found by plugging it in to the USB port, and opening the `Garmin/Activities` folder that appears on the USB device.

Here's an example GPS file, from a ride to Mt Beauty in Victoria, Australia:
```{r}
library(fit)
data <- read.fit('examples/mt_beauty.fit')
```
The names of the available data tables can be found using `names()`. 
```{r}
names(data)
```
My Garmin records exercise progress as `record` messages every few seconds, so if we want to analyse a ride we'd be interested in the `record` table:
```{r}
head(data$record)
```

The table contais data like GPS coordinates (latitude, longitude, altitude), cadence, heart rate, speed 
and distance travelled, depending on which gadgets you have attached to your device. If you have a power
meter, that data will show up here, too. 

The units for each field are available in the table's 'units' attribute. 
```{r}
attr(data$record,'units')
```

The timestamp, given in seconds, is the time elapsed since Jan 1, 1990. You might want to subtract the first value in the table, which would give you the number of seconds elapsed in your session.

Example analysis
----------------
Here's a plot of my elevation, in meters, as a function of the number of minutes I was riding. (It was a fairly decent hill!)
```{r}
library(ggplot2)
pdata <- with(data$record, data.frame(alt = altitude, time = (timestamp-timestamp[1])/60))
ggplot(pdata, aes(y=alt, x=time)) + geom_line() +
  ggtitle("Elevation vs Time") + xlab("time (minutes)") + ylab("elevation (m)")
```

Now let's try to answer a more interesting question: how much harder does my heart work when I'm riding uphill?

We can compute the average gradient using the familiar formula, `gradient = change in elevation/distance travelled`. 
The elevation readings are pretty noisy and only measured in whole meters, so I'll smooth the gradients by averaging over 10 observations, which is about 30-40 seconds of riding. (We can actually do better than this, using a Kalman filter or similar, but that's overkill for what we need now.)

As we might expect, the relationship is upward sloping: the steeper the hill, the faster my heart is working. Travelling down hill (negative gradient) is the easiest, and my heart is usually beating 150-170 beats per minute. But uphill, my heart can be doing anything from 170 to 205 or so when it's really steep. (Ouch!) 

And what's really interesting is that the relationship is non-linear: as the road gets steeper, the increase in heart rate diminishes. This is probably because there's only so hard my heart can actually beat---and rather than suffer cardiac arrest, I slow the bike down a bit and let my heart muscle rest.
```{r}
pdata <- data$record[-(1:10),c("heart_rate","timestamp")]
# compute average gradient, as %
pdata$gradient <- with(data$record, 100 * diff(altitude,lag=10) / diff(distance,lag=10))
pdata <- subset(pdata, complete.cases(pdata) & abs(gradient) < 7.5 & gradient != 0) # drop outliers
ggplot(pdata, aes(x=gradient, y=heart_rate)) + 
  geom_point(alpha=0.5) + geom_jitter() +
  stat_smooth(method="lm", formula=y ~ poly(x, 2)) +
  ggtitle("Heart rate vs gradient")
```

Making maps
-----------
Since we have GPS data, it'd be a shame not to plot the data on a map. One way to do this is to use David Kahle and Hadley Wickham's excellent `ggmap` package.

In the map below, we get a map image from Google Maps, and overlay the path I travelled, using my heart rate again for the colour.
```{r fig.width=8, warning=FALSE, message=FALSE}
library(ggmap)

garmin.data <- read.fit('examples/mt_beauty.fit')
points <- subset(garmin.data$record, complete.cases(garmin.data$record))

map <- get_googlemap(center = colMeans(points[,c('position_long','position_lat')]),
                     zoom = 12, maptype = c("hybrid"))

ggmap(map) +
  geom_path(aes(x = position_long, y = position_lat, colour = heart_rate),
             data = points, size = 2) +
  scale_colour_gradientn(colours = rainbow(3)) +
  ggtitle("Mt Beauty, 23 January 2014")
```

The GPS coordinates are suprisingly accurate. We discussed above that the altitude data is fairly noisy, although this is mostly because altitudes are given as integers, and there isn't much vertical movement when you ride. This means that round-off error is large compared to actual vertical movement, and so contributes a lot to the signal-to-noise ratio. But the opposite seems to be true for the longitude and latitude: the error is very small compared to actual horizontal movement.

For a visual illustration, check out this excerpt of my daily commute, which runs along the bike path on the western side of the Sydney Harbour Bridge. The GPS path tracks the bike lane accurately the whole way:
```{r fig.width=8, warning=FALSE, message=FALSE}
load('examples/commute.rda') # subset of a daily commute (R data format)
map <- get_googlemap(center = c(151.210787,-33.852306), zoom = 16, maptype = c("satellite")) # Harbour Bridge
ggmap(map, extent = "device") +
  geom_path(aes(x = position_long, y = position_lat),
             data = commute, size = 1, colour = 'blue')
```

But now look at the measured altitude over the same distance. The Bridge is cambered, so that its center is higher than the edges, but the road certainly isn't as jagged as the chart below suggests. (The steep drop at 2km marks the stairs at the North Sydney end.) In reality, the bike would have travelled a smooth path, passing between the measured points. So the vertical error, although probably not all that big in absolute terms, is certainly large relative to the total vertical distance travelled in a ride.
```{r, fig.width=8}
ggplot(commute, aes(y=altitude, x=distance)) + 
  geom_path() + ggtitle("Harbour Bridge bike lane: measured elevation") +
  xlab("distance (m)") + ylab("altitude (m)")
```

Have fun.
